---
title: "wastewater_viromics_report.Rmd"
output: html_notebook
---

### Libraries

```{r}
library(dplyr)
library(stringr)
library(lubridate)
library(readr)
```

# 1. Read the HTP-filtered CSV generated by the bash efetch script

```{r}
htp <- read_csv(
  "metadata/PRJNA729801_htp_.filtered.csv"
)
```

# htp has columns like:
# Run, BioSample, SampleName, LibraryName, LibraryStrategy,
# LibrarySource, LibraryLayout, Platform, Model, spots, bases,
# avgLength, size_MB, download_path
# [file:121]

# 2. Parse date from SampleName, e.g. "HTP_11_3_2020_INF_unenriched"

```{r}
parse_htp_date <- function(sample_name) {
  # expect pattern HTP_mm_dd_yyyy
  m <- str_match(sample_name, "HTP_(\\d{1,2})_(\\d{1,2})_(\\d{4})")
  if (any(is.na(m))) return(NA_date_)
  mm <- as.integer(m[2])
  dd <- as.integer(m[3])
  yyyy <- as.integer(m[4])
  as_date(sprintf("%04d-%02d-%02d", yyyy, mm, dd))
}
```

# 3. Infer treatment from SampleName / LibraryStrategy

```{r}
infer_treatment <- function(sample_name, library_strategy) {
  if (str_detect(sample_name, "INF_unenriched")) {
    "unenriched"
  } else if (library_strategy == "Targeted-Capture") {
    "enriched"
  } else {
    "unknown"
  }
}
```

# 4. Build metadata with parsed date, treatment, site, matrix

```{r}
htp_meta <- htp %>%
  mutate(
    date = parse_htp_date(SampleName),
    treatment = infer_treatment(SampleName, LibraryStrategy),
    site = "HTP",
    matrix = if_else(str_detect(SampleName, "INF"), "INF", "unknown")
  )
```

# 5. Restrict to the five paired dates we’re using

```{r}

target_dates <- as_date(c(
  "2020-11-03",
  "2020-11-24",
  "2020-12-04",
  "2020-12-22",
  "2021-01-03"
))


htp_pairs <- htp_meta %>%
  filter(
    date %in% target_dates,
    treatment %in% c("unenriched", "enriched")
  )
```


# 6. Construct sample_id and pick the 10 useful columns

```{r}
samples_htp <- htp_pairs %>%
  mutate(
    sample_id = paste0("HTP_", date, "_", treatment)
  ) %>%
  transmute(
    sample_id,               # used by Snakemake rules
    site,                    # HTP
    date = as.character(date),  # ISO date for plotting
    treatment,               # enriched vs unenriched
    matrix,                  # INF vs unknown
    srr = Run,               # SRR accession
    biosample = BioSample,   # SAMN accession
    library_strategy = LibraryStrategy,
    library_source = LibrarySource,
    layout = LibraryLayout
  ) %>%
  arrange(date, treatment)
```

# 7. Write the curated sample sheet

```{r}

write_csv(
  samples_htp,
  "config/samples_htp.csv"
)
```




